% !TeX root = main.tex
\section{The electric field in material media; dielectric polarisation, conductivity, mobility of charge carriers}
    \label{sec:electric-field}\todo[backgroundcolor=blue, textcolor=white, linecolor = blue]{content 2 starts here}

    \subsection{Polarisation}

        Dielectric polarisation is the phenomenon consisting in the formation of electric dipoles, or in the orientation of existing ones, under the influence of an external electric field. As a result of polarisation an internal electric field arises which partially balances the external field. In the linear approximation the polarisation is proportional to the field, $\vec{P} = \varepsilon_0\chi\vec{E}$ (\cref{eq:polarisation-vector}), with $\chi$ the electric susceptibility.
        % NOTE (dedup): folded the standalone P = eps0 chi E display into the sentence and
        % cross-referenced \cref{eq:polarisation-vector} in content_1 instead of repeating it.

    \paragraph{Electronic polarisation}

        Electronic polarisation is the result of the displacement, under the influence of an electric field, of the negatively charged electron clouds with respect to the positive nuclei. It occurs in all materials. It is a very fast process, and therefore, when investigated with an alternating electric field, the effects associated with it are observed up to frequencies corresponding to the ultraviolet. The dipole moment $p_e$ is determined by the electronic polarisability $\alpha_e$:

        \begin{equation*}
            p_e = \alpha_e \cdot E
        \end{equation*}

    \paragraph{Ionic polarisation}

        Ionic polarisation occurs only in materials with ionic bonds. It is the result of the displacement of ions in the crystal lattice of the material---those of one sign in one direction, those of the other sign in the other. It is the slowest of all the polarisation processes, since it requires the movement of many atoms bound in the crystal lattice.

    \paragraph{Orientational polarisation}

        Orientational polarisation, also called dipolar polarisation, occurs only in polar dielectrics, that is, those whose molecules form permanent dipoles (they possess their own dipole moment). In an electric field an ordering torque acts on them. At the same time the ordering is destroyed by thermal vibrations, which causes orientational polarisation to be temperature dependent.

    \subsection{Conductivity and mobility}

        Conduction is the phenomenon of the flow of electric charges occurring in a material under the influence of an electric field. The quantity characterising electrical conduction is the conductivity $\sigma$:

        \begin{equation*}
            \sigma = q n \mu
        \end{equation*}

        where $q$ is the charge of the carriers, $n$ their concentration and $\mu$ the carrier mobility. The mobility of the carriers depends on the applied field and on the drift velocity. Temperature is one of the factors affecting conductivity. A change in temperature may affect a change both in the concentration and in the mobility of the carriers. The most common types of conduction are: electronic, where the charge carriers are electrons, hole conduction, where the charge carriers are electrons moving in the valence band, and ionic, where the charge carriers are ions.

\section{The magnetic field in material media; classification of magnetic media, microscopic explanations of the exchange interaction between the magnetic moments of the constituents (atoms, ions, molecules)}
\label{sec:magentic-field}

    On the microscopic scale every medium exhibits magnetic properties, on account of the spin of its constituent particles and of the motion of electrons about the atomic nuclei. On the macroscopic scale, however, magnetic phenomena are by no means so common, which follows from the absence of long-range order and from destructive interference between the constituent magnetic moments. Long-range order may nevertheless be introduced as a result of an external magnetic field $B$, similarly to the case of inducing polarisation of a medium with an electric field. In general, when describing magnetic fields in material media we use constitutive relations analogous to those for the electric field. Recalling the definition of the magnetic field strength $\vec{H} = \vec{B}/\mu_0 - \vec{M}$ (\cref{eq:H-def}), and writing the magnetisation as linear in $\vec{H}$:

    \begin{align*}
        \vec{M} &= \chi_m\vec{H} \\
        \vec{H} &= \frac{1}{\mu_0(1+\chi_m)}\vec{B}
    \end{align*}

    where $\vec{H}$ is the magnetic field strength, $\vec{M}$ is the magnetisation of the medium and $\chi_m$ the magnetic susceptibility. Interactions with a magnetic field are, however, more nuanced than in the case of the electric field. According to the Bohr--van Leeuwen theorem, a classical system will always have zero mean magnetisation, and therefore any magnetisation must be the result of quantum effects. Three kinds of medium are usually distinguished.

    \subsection{Paramagnets}
    \todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: defined the Curie constant and its temperature dependence and ``Fermi gas'' in plain terms; added material examples.}

        The most common group. They have positive susceptibility, so the induced magnetisation points along the applied field. It comes from unpaired electrons, whose magnetic moments turn to line up with an external field. Common paramagnets are aluminium, platinum and oxygen. Two standard models:

        \begin{itemize}
            \item Curie paramagnetism --- a set of independent atomic moments that do not interact with one another, as in the paramagnetic salts of transition-metal or rare-earth ions. Balancing their alignment against the randomising effect of heat gives Curie's law for the susceptibility,
            \begin{equation*}
                \chi_m = \frac{C}{T},
            \end{equation*}
            where $T$ is the absolute temperature and $C$ (the Curie constant) is fixed by the size and number of the atomic moments. The susceptibility falls as the temperature rises because thermal agitation fights the alignment.
            \item Pauli paramagnetism --- in a metal the conduction electrons (as in an ordinary metal such as sodium) are treated as a \emph{Fermi gas}: a gas of electrons that barely interact, but which (because no two electrons may share the same quantum state) fill the available energy levels from the bottom up to a top level called the Fermi level. Only the few electrons near that top level are free to reorient in a field, so the resulting paramagnetism is weak and almost independent of temperature.
        \end{itemize}

    \subsection{Diamagnets}
    \todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: clarified ``all electrons paired'' and how bound electrons respond in an insulator (Larmor precession, not free circulation); added material examples.}

        Diamagnetism is the appearance of a magnetisation \emph{opposite} to the applied field, i.e.\ a negative susceptibility. Every material has this response, but it is very weak, so we call a material a diamagnet only when it is the dominant effect. That happens when the atoms carry no permanent magnetic moment---all electron spins paired---so there is no paramagnetic contribution to mask it. Typical diamagnets are bismuth (the strongest ordinary one), copper, gold and water.

        A special case is that of superconductors, which expel the magnetic field from their interior completely through the Meissner effect. The remaining diamagnets are described by:

        \begin{itemize}
            \item Langevin diamagnetism --- describes insulators (such as water and the inert gases), where the electrons are bound in atoms rather than free to conduct. An applied field makes each bound electron orbit precess (Larmor precession, \cref{subsec:quantum-precession}). A precessing orbit is effectively a small current loop, and the magnetic moment it produces points against the applied field, which is the origin of the diamagnetic response.
            \item Landau diamagnetism --- treats the same Fermi gas of conduction electrons in a metal as Pauli paramagnetism, but now accounts for the field bending the electrons' orbital motion rather than flipping their spin. This orbital effect is diamagnetic.
        \end{itemize}

    \subsection{Ferromagnets}
    \todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: defined remanence, gave one-line meanings of the Ising/XY/Heisenberg models, and stated the macroscopic effect of antiferro-/ferrimagnets; added material examples.}

        A third category appears once we include the \emph{exchange interaction}---a quantum effect that makes neighbouring moments prefer to point the same way as each other, not just as the external field. Regions of aligned moments (magnetic domains) form, giving a strong susceptibility and \emph{remanent} magnetisation: magnetisation that survives after the external field is switched off, which is what makes a permanent magnet. The classic ferromagnets are iron, cobalt and nickel.

        Such materials are analysed with statistical-physics lattice models that differ in how freely each moment may point: the Ising model allows only up or down, the XY model allows any direction in a plane, and the Heisenberg model allows any direction in space (the most general).

        Closely related are antiferromagnets and ferrimagnets, in which neighbouring moments point in \emph{opposite} directions. In an antiferromagnet the opposing moments are equal and cancel, so there is almost no net magnetisation (manganese oxide and chromium are examples). In a ferrimagnet they are opposite but unequal, so the cancellation is only partial and a net magnetisation remains---macroscopically it behaves like a weaker ferromagnet (magnetite is the classic example).

\section{Electromagnetic waves as examples of solutions of Maxwell's equations; polarisation, reflection and refraction of waves, the Doppler effect}

\subsection{Electromagnetic waves as examples of solutions of Maxwell's equations}

Let us recall the general Maxwell equations in vacuum (\cref{eq:maxwell-vacuum}):

\begin{equation*}
    \begin{aligned}
        \nabla \cdot \vec{E} &= 0\\
        \nabla \cdot \vec{B} &= 0\\
        \nabla\times\vec{E} &= -\frac{\partial \vec{B}}{\partial t}\\
        \nabla\times\vec{B} &= \mu_{0}\epsilon_{0}\frac{\partial\vec{E}}{\partial t}
    \end{aligned}
\end{equation*}

We shall find an electromagnetic wave  is a solution of this system of equations. To do it, take the curl of the third equation:

\begin{equation*}
    \nabla\times\left(\nabla\times\vec{E}\right) = \nabla\times\left(- \frac{\partial\vec{B}}{\partial t}\right).
\end{equation*}

We make use of the following identity:

\begin{equation*}
    \nabla\times(\nabla\times\vec{F}) = \nabla\cdot(\nabla\cdot\vec{F}) - \nabla^{2}\cdot\vec{F}.
\end{equation*}

Then, we have:

\begin{equation*}
    \begin{aligned}
        \mathrm{LHS} = \nabla\times\left(\nabla\times\vec{E}\right)                 &= \nabla\cdot\left(\nabla\cdot\vec{E}\right) - \nabla^{2}\cdot\vec{E} =  \nabla\cdot\left( 0 \right) - \nabla^{2}\cdot\vec{E} = - \Delta \vec{E}\\
        \mathrm{RHS} = \nabla\times\left(- \frac{\partial\vec{B}}{\partial t}\right) &= - \frac{\partial}{\partial t}\left(\nabla\times\vec{B}\right) = - \frac{\partial}{\partial t} \left( \mu_{0}\epsilon_{0}\frac{\partial\vec{E}}{\partial t} \right) = - \mu_{0}\epsilon_{0}\frac{\partial^{2}\vec{E}}{\partial t^{2}},
    \end{aligned}
\end{equation*}

where, for the LHS, we have used the identity and the first Maxwell equation, and for the RHS, we have used the fourth Maxwell equation and  

\begin{equation}
    \Delta \vec{F} = \nabla^{2} \vec{F} = \left( \frac{\partial^2}{\partial x^2} + \frac{\partial^2}{\partial y^2} + \frac{\partial^2}{\partial z^2} \right) \vec{F}
\end{equation}

is the Laplacian operator.

Equating both sides we obtain:

\begin{equation}\label{eq:em-wave}
    \left(\Delta - \mu_{0}\epsilon_{0}\frac{\partial^{2}}{\partial t^{2}}\right)\vec{E} = 0
\end{equation}

We proceed analogously with the fourth Maxwell equation and find the equation for $\vec{B}$.

\begin{equation*}
    \left(\Delta - \mu_{0}\epsilon_{0}\frac{\partial^{2}}{\partial t^{2}}\right)\vec{B} = 0
\end{equation*}

Both of these equations are standard forms of the wave equation, generally written as:

\begin{equation*}
    \left(\Delta - \frac{1}{v^{2}}\frac{\partial^{2}}{\partial t^{2}}\right)\vec{F} = 0,
\end{equation*}

where $v$ is the propagation speed. From this it follows that the speed of propagation of an electromagnetic wave in vacuum is 

\begin{equation}
    v = c = \frac{1}{\sqrt{\mu_{0}\epsilon_{0}}}.
\end{equation}

Performing an analagous derivation for the Maxwell equations in a medium, we find:

\begin{equation}
    \left(\Delta - \mu_{0}\epsilon_{0}\frac{\partial^{2}}{\partial t^{2}}\right)\vec{E} = \overbrace{\nabla\left( \frac{\rho}{\epsilon_{0}} \right) + \underbrace{\mu_{0}\frac{\partial\vec{J_{\text{f}}}}{\partial t}}_{\substack{\text{Important in} \\ \text{conductors}}} + \underbrace{\mu_0\frac{\partial^2 \vec{P}}{\partial t^2}}_{\substack{\text{Important in} \\ \text{insulators}}}}^{\text{0 in vacuum}}
\end{equation}

\paragraph{Solution to the wave equation}
\todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: noted the physical field is the real part and that the direction of $\vec{E_0}$ sets the polarisation, leading into the next subsection.}

The wave equation is solved by a plane wave of the form

\begin{equation}
    \vec{E}(\vec{r},t) = \vec{E_0}\,e^{i\left(\vec{k}\cdot\vec{r} - \omega t + \phi\right)},
\end{equation}

with the physical field understood as the real part, using $e^{i\alpha} = \cos{\alpha} + i\sin{\alpha}$. Here $\vec{k}$ points along the direction of travel, and for this transverse wave the constant vector $\vec{E_0}$ is perpendicular to $\vec{k}$. The magnetic field $\vec{B}$ is perpendicular to both, with $|\vec{B}| = |\vec{E}|/c$. The direction in which $\vec{E_0}$ points is called the \emph{polarisation} of the wave, discussed next.

\subsection{Polarisation of a wave}
\todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: gave concrete linear/circular/elliptical cases tied to the plane-wave solution, with a light mention of the Jones vector.}

Polarisation describes the pattern that the electric-field vector traces out, in the plane perpendicular to the direction of propagation, as the wave passes. Only transverse waves can be polarised, since the oscillation must have a choice of directions perpendicular to the travel direction (a longitudinal wave such as sound has none). Taking propagation along $z$, the field lies in the $x$--$y$ plane and the main cases are:

\begin{itemize}
    \item linear --- the two perpendicular components oscillate in phase, so $\vec{E}$ stays along a fixed line.
    \item circular --- the components have equal amplitude but are a quarter-period ($90^\circ/\frac{\pi}{2}$) out of phase, so the tip of $\vec{E}$ traces a circle.
    \item elliptical --- the general case, tracing an ellipse.
\end{itemize}

The two complex components $(E_x, E_y)$ of the plane wave above therefore fix the polarisation through their relative amplitude and phase. This pair is known as the Jones vector.

\subsection{The laws of reflection and refraction of electromagnetic waves}
\todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: defined ``ray'' and the s/p polarisations, and stated what Brewster's angle does to them (Fresnel coefficients named, forms omitted).}

A \emph{ray} is a line drawn along the direction in which the wave (and its energy) travels, at right angles to the wavefronts. The laws below are stated in terms of rays.

\textbf{The law of reflection.} The angle of reflection equals the angle of incidence, and the incident ray, the reflected ray and the normal to the surface lie in one plane. Reflection changes only the direction of propagation, not the wavelength.

\textbf{The law of refraction (Snell's law).} The incident and refracted rays and the normal lie in one plane, and the angles satisfy

\begin{equation}\label{eq:snell}
    n_1 \sin \theta_1 = n_2 \sin \theta_2,
\end{equation}

where $n_1$, $n_2$ are the refractive indices of the two media, $\theta_1$ the angle of incidence and $\theta_2$ the angle of refraction, both measured from the normal.

To describe what happens to the polarisation on reflection, we fix the \emph{plane of incidence}: the plane containing the incident ray and the normal. A wave is called $p$-polarised (parallel) when its electric field lies in this plane, and $s$-polarised (perpendicular/senkrecht) when its field is perpendicular to it. The fractions of amplitude reflected and transmitted differ for the two, and are given by the Fresnel coefficients (forms omitted here).

\textbf{The Brewster angle} $\alpha_B$ is the angle of incidence at which the reflected and refracted rays are perpendicular to one another. At this angle the reflected $p$-component vanishes, so the reflected light is completely $s$-polarised (perpendicular to the plane of incidence). It is given by

\begin{equation}\label{eq:brewster}
    \tan{\alpha_{B}} = \frac{n_{2}}{n_{1}},
\end{equation}

with the light incident from the side of medium $n_1$ (taking $n_1 < n_2$).

\subsection{The Doppler effect}
\todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: replaced the convoluted derivations with standard $\pm$ formulae, split acoustic vs.\ relativistic EM, and cross-referenced $\beta$ instead of restating it.}

\textbf{The Doppler effect} is the difference between the frequency a source emits and the frequency an observer records when the two move relative to one another. For waves that travel in a medium (such as sound) the speeds of source and observer are measured relative to that medium, and the two cases are not symmetric. For light in vacuum, which needs no medium, only the relative velocity matters.

\paragraph{Acoustic (mechanical) waves} With $v$ the wave speed in the medium, $v_o$ the observer's speed and $v_s$ the source's speed (both relative to the medium),

\begin{equation}\label{eq:doppler-acoustic}
    f' = f\,\frac{v \pm v_o}{v \mp v_s},
\end{equation}

where the upper signs apply when that party moves \emph{towards} the other (raising the pitch) and the lower signs when it moves away. Observer and source enter differently: the observer changes the wave speed it meets, while the source changes the wavelength it emits.

\paragraph{Electromagnetic waves} Here the effect is relativistic and depends only on the relative speed through $\beta = v/c$ (\cref{eq:gamma}). For source and observer separating along the line of sight,

\begin{equation*}
    f' = f\,\sqrt{\frac{1-\beta}{1+\beta}} \qquad (\text{redshift}),
\end{equation*}

and swapping the sign of $\beta$ gives the approaching (blueshift) case. If the relative velocity makes an angle $\phi$ with the line of sight,

\begin{equation}\label{eq:doppler-em}
    f' = f\,\frac{\sqrt{1-\beta^{2}}}{1+\beta\cos{\phi}}.
\end{equation}

At $\phi = \tfrac{\pi}{2}$ this leaves the \emph{transverse} Doppler effect, $f' = f\sqrt{1-\beta^{2}} = f/\gamma$, a pure redshift caused by time dilation that has no counterpart for mechanical waves.

\section{Mechanical waves, electromagnetic waves, matter waves; similarities and differences}
\todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten for consistency: cross-referenced the wave equation and the mechanical-waves description (removing the repeated paragraph), and labelled the de Broglie relation.}

\textbf{Electromagnetic waves} are oscillating electric and magnetic fields that satisfy the wave equation derived earlier (\cref{eq:em-wave}). They are transverse, travel at $c$ in vacuum and need no medium. In the quantum picture the wave is a stream of photons, each with energy $E = \hbar\omega$ and momentum $p = \hbar k$, where $\omega$ is the angular frequency and $k$ the wave number.

\textbf{Mechanical waves} are disturbances passing through an elastic medium, as described in \cref{sec:mechanical-media}. They require a medium and may be longitudinal or transverse, with a speed set by the properties of the medium.

\textbf{Matter waves} express the fact that every particle also behaves as a wave. The associated (de Broglie) wavelength is

\begin{equation}\label{eq:de-broglie}
    \lambda = \frac{h}{p} = \frac{h}{mv} \quad(\text{non-relativistically}),
\end{equation}

so the wave nature is easiest to see for light particles such as electrons, whose diffraction and interference are used in electron microscopy and diffraction imaging. A localised particle corresponds not to a single wavelength but to a \emph{wave packet}---a superposition of many waves. The same relations $E = \hbar\omega$ and $p = \hbar k$ apply.

\textbf{Common features and differences.} All three show interference and diffraction, obey the laws of reflection and refraction, and can form standing waves by superposition. They differ in what they need and what they are: mechanical waves need a medium, whereas electromagnetic and matter waves do not. Electromagnetic and matter waves are quantum in nature---for a matter wave the squared amplitude gives the probability of finding the particle, while for light it corresponds to the number of photons.

\section{Phase transitions; kinds of phase transition, the order parameter of the transition, conditions for the coexistence of phases (examples of phenomena)}
\todo[backgroundcolor=green!25,linecolor=green!55!black]{Full rewrite: fixed the reversed entropy claim for the order parameter, identified the potential as the Gibbs free energy (not enthalpy), and simplified the terminology with concrete examples.}

A \emph{phase} is a region that is uniform in its physical and chemical properties---ice, liquid water and steam are three phases of the same substance. A \emph{phase transition} is an abrupt change in these properties when a control variable such as temperature or pressure is varied. Transitions may involve the state of matter, the crystal structure, magnetic ordering or superconductivity.

To describe them we use a thermodynamic potential---for changes at fixed temperature and pressure this is the \emph{Gibbs free energy}, the quantity a system minimises at equilibrium (it is not the enthalpy). It is written as a function of an \emph{order parameter}: a quantity that is zero in the more symmetric, higher-entropy (disordered) phase and becomes nonzero in the ordered phase. Two examples are the magnetisation of a ferromagnet (zero above the Curie temperature, nonzero below) and the density difference between coexisting liquid and gas.

The \emph{Ehrenfest classification} sorts transitions by how the chemical potential $\mu$ (the free energy per particle) behaves. A transition is of the $n$-th kind when the $n$-th derivative $\partial^{n}\mu/\partial T^{n}$ is the first that is discontinuous.

\begin{itemize}
    \item First order --- a first derivative jumps, producing latent heat and a sudden change of density. Melting and boiling are examples.
    \item Second order (continuous) --- the first derivatives are continuous but a second derivative jumps or diverges, with no latent heat. A ferromagnet losing its magnetisation at the Curie temperature is the standard example.
    \item Third and higher orders essentially do not occur in practice. Continuous transitions such as the onset of superfluidity also belong here.
\end{itemize}

The conditions for phases to coexist are given by the \emph{Gibbs phase rule},

\begin{equation}\label{eq:gibbs-phase}
    f = s - r + 2,
\end{equation}

where $s$ is the number of chemical components, $r$ the number of coexisting phases and $f$ the number of intensive variables (such as temperature and pressure) that may be changed while keeping those phases in equilibrium. It requires the system to be in thermal, mechanical and chemical equilibrium. For a single substance ($s = 1$):

\begin{itemize}
    \item $r = 1$: one phase, $f = 2$ --- an area on the pressure--temperature diagram.
    \item $r = 2$: two phases coexisting, $f = 1$ --- a coexistence curve.
    \item $r = 3$: three phases, $f = 0$ --- the triple point.
\end{itemize}

\section{Experimental foundations of quantum mechanics; examples of phenomena that can be explained in quantum mechanics, but not in classical theory}
\todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: tightened all four effects, fixed typos, added the photoelectric and Compton relations, and cross-referenced the Planck distribution.}

    \subsection{The photoelectric effect}

        The photoelectric effect is the emission of electrons from a material illuminated by electromagnetic radiation. Classically the energy of the emitted electrons should grow with the intensity of the light. Experimentally, however, the kinetic energy of each electron depends only on the \emph{frequency} of the light, while the intensity sets only the \emph{number} of electrons. This is explained if light arrives in quanta (photons) of energy $E = h\nu$: one photon frees one electron, giving it an energy $h\nu$ less the work function $W$ (the energy binding the electron to the material),

        \begin{equation*}
            E_\text{kin} = h\nu - W.
        \end{equation*}

        Below a threshold frequency no electrons are emitted at all, because a single photon then lacks the energy to free one.

    \subsection{The Compton effect}

        The Compton effect is the scattering of X-ray or gamma radiation on free or weakly bound electrons, in which the scattered radiation emerges with a longer wavelength. It is explained by treating the photon as a particle carrying both energy and momentum: in the collision it hands some of each to the electron, so the scattered photon has less energy and hence a longer wavelength. The shift grows with the scattering angle $\theta$,

        \begin{equation*}
            \Delta\lambda = \frac{h}{m_e c}\left(1 - \cos\theta\right),
        \end{equation*}

        and has no classical explanation, since a classical wave would scatter without changing its wavelength.

    \subsection{Young's experiment}

        Coherent light passed through two closely spaced slits produces bright and dark interference fringes on a screen, exactly as expected for waves. The quantum surprise is that particles such as electrons, fired one at a time at a double slit, gradually build up the same interference pattern---yet each arrives as a single discrete impact at an unpredictable point. The wave nature governs where each particle is \emph{likely} to land, but the individual outcome is inherently probabilistic, which classical mechanics cannot account for.

    \subsection{Quantisation of the energy levels in the hydrogen atom}

        Atomic spectra show that the hydrogen atom can occupy only a discrete set of energy levels, whereas classical physics allows a continuous range of orbits. Classical physics fails here in two ways: it cannot explain the sharp spectral lines, and it cannot even keep the atom stable---a classically orbiting electron is an accelerating charge, so it would radiate its energy away and spiral into the nucleus almost at once. Quantisation of the allowed energies removes both problems.

    \subsection{The ultraviolet catastrophe}

        Classical physics (the Rayleigh--Jeans law) predicted that a black body in thermal equilibrium would radiate ever more strongly at shorter wavelengths, so that the total emitted power would diverge to infinity in the ultraviolet---plainly at odds with experiment. In 1900 Planck removed the divergence by assuming that radiation is emitted and absorbed only in discrete packets of energy $E = h\nu$. This suppresses the high-frequency modes and yields the correct spectrum, the Planck distribution (\cref{eq:planck}).

\section{The uncertainty principle and the limits of measurement accuracy}
\label{sec:uncertainty}
\todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: corrected the Robertson relation (squared right-hand side), added a plain-language commutator introduction and the sine-wave/localised-spike example.}

    The Heisenberg uncertainty principle states that certain pairs of observables cannot both have a sharp value in the same state: the more precisely one is fixed, the less precisely the other can be. This is a fundamental feature of quantum states, not a shortcoming of our instruments.

    Whether two observables $\hat{A}$ and $\hat{B}$ can be sharp together is decided by their \emph{commutator}

    \begin{equation*}
        \comm{\hat{A}}{\hat{B}} = \hat{A}\hat{B} - \hat{B}\hat{A},
    \end{equation*}

    which measures how much the order of the two operations matters. If it vanishes the observables commute and can be measured simultaneously to any accuracy. If it does not, the spreads of the two quantities obey the Robertson relation

    \begin{equation}\label{eq:uncertainty}
        \Delta^2 \hat{A}\,\Delta^2 \hat{B} \geq \frac{1}{4}\left| \ev{\comm{\hat{A}}{\hat{B}}} \right|^2,
    \end{equation}

    where the variance $\Delta^2 \hat{A} = \bra{\psi}\hat{A}^2\ket{\psi} - \left|\bra{\psi}\hat{A}\ket{\psi}\right|^2$ measures the spread of $\hat{A}$ about its mean.

    The most important case is position and momentum. Since $\comm{\hat{x}}{\hat{p}_x} = i\hbar$, the relation gives

    \begin{equation*}
        \Delta^2 x\,\Delta^2 p_x \geq \frac{\hbar^2}{4},
    \end{equation*}

    saturated (turned into an equality) by a Gaussian wave packet, which is therefore the best-localised state possible in position and momentum at once.

    The trade-off is easiest to see for a wave. A pure sine wave has a single wavelength, so by the de Broglie relation ($p = \hbar k$, \cref{eq:de-broglie}) its momentum is perfectly defined---but it extends over all space, so its position is completely undefined. A sharply localised spike has a definite position, but building it requires adding together waves of every wavelength, so its momentum is completely undefined. Any real state is a compromise between these extremes.

    This should not be confused with ``Heisenberg's microscope'', the argument that pinning down a particle's position needs light of ever shorter wavelength, whose photons then carry ever larger momentum and so disturb the particle more. That picture is misleading, because it still assumes the particle has a definite position waiting to be measured, which is exactly what the principle denies.

\section{Approximate methods of quantum mechanics; perturbation theory for eigenvalues, Fermi's golden rule, the Born approximation for scattering amplitudes}
\label{sec:approximate-methods}

    Quantum mechanics describes the behaviour of systems at the atomic and subatomic level. However, it rests to a large extent on solving differential equations which we are able to solve analytically only in relatively simple cases. Sometimes even determining the Hamiltonian itself is difficult, and therefore for more complex systems approximate methods are used, which make it possible to obtain useful results.

    \subsection{Stationary perturbation theory}

    Given a known Hamiltonian $H_0$, for example that of the hydrogen atom, we may wish to introduce a slight modification into the system, for example a weak external electric field. Solving it analytically again would be just as highly non-trivial as the hydrogen atom itself. We may instead treat our correction as a perturbation in the form of an additional term $\lambda H'$. The method itself is the quantum counterpart of an expansion in a Taylor series with respect to the parameter $\lambda$ and of calculating successive corrections. The smaller the perturbation, the fewer corrections need to be calculated, and often one or two suffice. They are given by the following formulae:

    % NOTE: the source wrote the kets on the right as \ket{n^{(0)}}; changed to
    % \ket{\psi_n^{(0)}} for consistency with the bra notation.
    \begin{align}
        E_n^{(1)} &= \bra{\psi_n^{(0)}} H' \ket{\psi_n^{(0)}} \label{eq:pt-e1}\\
        \psi_n^{(1)} &= \sum_{m \neq n} \frac{\bra{\psi_m^{(0)}} H' \ket{\psi_n^{(0)}}}{E_n^{(0)} - E_m^{(0)}} \psi_m^{(0)} \label{eq:pt-psi1}\\
        E_n^{(2)} &= \sum_{m \neq n} \frac{|\bra{\psi_m^{(0)}} H' \ket{\psi_n^{(0)}}|^2}{E_n^{(0)} - E_m^{(0)}} \label{eq:pt-e2}
    \end{align}

    \subsection{Fermi's golden rule}

    The method of stationary perturbation theory described above assumes that the perturbing potential is invariant in time. There also exists a dynamical method, but it is far more complicated. In the special case, however, in which the perturbing potential varies periodically in time, one may apply Fermi's golden rule, an important result of that calculation. This solution gives the probability per unit time of transitions between two quantum states $\psi_n^{(0)}$ and $\psi_m^{(0)}$, for a perturbation $H'e^{-i\omega t}$:

    \begin{equation}\label{eq:fermi-golden-rule}
    \Gamma_{n\rightarrow{}m} \approx \frac{2\pi}{\hbar} |\bra{\psi_m^{(0)}}H'\ket{\psi_n^{(0)}}|^2 \delta\left(E_n^{(0)} - E_m^{(0)} - \hbar\omega\right)
    \end{equation}

    \subsection{The Born approximation}
    \todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: fixed the scattering amplitude (factor $\hbar^2$ and minus sign), clarified the plane-wave set-up, defined the momentum transfer $K$, and distinguished it from the Born--Oppenheimer approximation.}

    A further application of perturbation theory is the Born approximation for scattering. A particle is scattered off a fixed, spherically symmetric potential $V(r)$. The incident particle is described by a plane wave, and far from the potential the state is this incident wave plus an outgoing spherical wave whose angular amplitude $f(\theta, \phi)$ is the \emph{scattering amplitude}. Its squared modulus $|f|^2$ is the differential cross-section, that is, how much is scattered into each direction.

    Writing the Schrödinger equation in an equivalent integral form (the Lippmann--Schwinger equation) and keeping only its first iteration gives the (first) Born approximation. For a spherically symmetric potential,

    \begin{equation}\label{eq:born}
    f(\theta) = -\frac{2m}{\hbar^2 K}\int_0^\infty r'V(r')\sin(Kr')\,dr',
    \end{equation}

    where $K = |\vec{k} - \vec{k}'| = 2k\sin(\theta/2)$ is the momentum transfer (the change in wave vector on scattering) and $k$ is the wave number. It works best for weak potentials or fast particles, where the scattering is a small perturbation. This is unrelated to the Born--Oppenheimer approximation of molecular physics, which separates the slow nuclear motion from the fast electronic motion.

\section{The Schrödinger equation as the basis for understanding the structure of the atom and of the chemical molecule; the periodic table of the elements, chemical bonds}

    \subsection{The Schrödinger equation}

        The Schrödinger equation governs how a quantum state evolves in time:

        \begin{equation}\label{eq:schrodinger}
            i\hbar\frac{d}{dt}\ket{\psi\left(t\right)} = \hat{H}\ket{\psi\left(t\right)}
        \end{equation}

        \todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: clarified the quantum numbers, the meaning of an orbital, and the binding energy (with the hydrogen value).}
        Solving it for an atom, physically acceptable solutions exist only for particular integer \emph{quantum numbers}: the principal $n$ (which sets the energy and the shell), the azimuthal $l$ (the magnitude of the orbital angular momentum) and the magnetic $m$ (its projection on the $z$ axis). Equivalently, the energy, the squared angular momentum and its $z$-component are quantised. Each solution $\Psi_{nlm}\left(r,\theta,\phi\right)$ is an \emph{orbital}---a spatial probability distribution for the electron. In chemistry the \textbf{electron shell} is the set of orbitals sharing the same principal quantum number $n$. Solving the equation also gives the \emph{binding energy}, the energy needed to remove the electron from the atom (for the hydrogen ground state, $13.6$~eV).

    \subsection{The periodic table of the elements}
    \todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: corrected the periodic law to depend on atomic number (not mass) and defined atomic number, mass number, period and group.}

        The \textbf{periodic table} arranges all the chemical elements in order of increasing \emph{atomic number} $Z$---the number of protons in the nucleus, which fixes the identity of the element and equals the number of electrons in the neutral atom. (The \emph{mass number} $A$, the total number of protons and neutrons, is a separate quantity and is not what orders the table.) The modern \textbf{periodic law} states that the properties of the elements vary periodically with atomic number. Mendeleev originally ordered the elements by atomic mass, which gives almost the same sequence, but the correct ordering is by atomic number.

        The layout of the table reflects the electron configuration:

        \begin{itemize}
            \item a \textbf{group} (vertical column) collects elements with a similar valence-shell configuration, and hence similar chemical behaviour.
            \item a \textbf{period} (horizontal row) corresponds to the principal quantum number $n$ of the outermost occupied shell.
        \end{itemize}

    \subsection{Different types of chemical bonds}
    \todo[backgroundcolor=green!25,linecolor=green!55!black]{Filled in: named the main bond types with one-line characteristics.}

        A chemical bond is an attraction that holds atoms together, arising in each case from the valence electrons rearranging to lower the energy. The main types are:

        \begin{itemize}
            \item \textbf{Ionic} --- one atom transfers electrons to another, and the resulting positive and negative ions attract electrostatically. It is strong and forms crystalline solids (e.g.\ NaCl).
            \item \textbf{Covalent} --- neighbouring atoms share pairs of electrons so that each attains a filled valence shell. It is strong and directional (e.g.\ H$_2$, diamond).
            \item \textbf{Metallic} --- positive ion cores sit in a shared ``sea'' of delocalised valence electrons, which accounts for the conductivity and malleability of metals (e.g.\ copper).
            \item \textbf{Hydrogen bonds and van der Waals forces} --- much weaker attractions. A hydrogen bond links a hydrogen atom to a nearby electronegative atom (it holds water together and binds the two strands of DNA), while van der Waals forces arise from weak, fluctuating dipoles.
        \end{itemize}

\section{Quantum tunnelling; examples}
\label{sec:tunnelling}
\todo[backgroundcolor=green!25,linecolor=green!55!black]{Rewritten: corrected the three-region wave function (oscillating outside, exponential inside) and the transmission coefficient (transmitted/incident, not reflected/incident).}

    Quantum tunnelling is the passage of a particle through a potential barrier higher than the particle's own energy---something forbidden in classical mechanics, where the particle simply lacks the energy to get over. It is possible because a quantum particle is described by a wave function whose squared modulus gives the probability of finding it at each point.

    Consider a particle of energy $E$ meeting a barrier of height $V_0 > E$. Solving the Schrödinger equation in the three regions gives:

    \begin{itemize}
        \item \textbf{before the barrier} --- an incident plus a reflected wave, $\psi(x) = Ae^{i k x} + Be^{-i k x}$.
        \item \textbf{inside the barrier} --- since $E < V_0$ the wave does not oscillate but decays exponentially, $\psi(x) = Ce^{-\kappa x} + De^{\kappa x}$, with $\kappa = \sqrt{2m(V_0 - E)}/\hbar$.
        \item \textbf{beyond the barrier} --- a transmitted wave travelling outwards, $\psi(x) = Fe^{i k x}$, with a smaller amplitude, $|F| < |A|$.
    \end{itemize}

    Because the exponential tail inside the barrier is still nonzero at the far edge, the transmitted amplitude does not vanish: there is a finite probability of the particle emerging. This is measured by the transmission coefficient

    \begin{equation*}
        T = \frac{|F|^2}{|A|^2},
    \end{equation*}

    which falls off exponentially with the width of the barrier and with $\sqrt{m(V_0 - E)}$.

    Nuclear fusion, the source of the Sun's energy, occurs to a large extent thanks to the tunnelling phenomenon. This phenomenon makes it possible to overcome the barrier of Coulomb repulsion of atomic nuclei at a temperature lower than would follow from the laws of thermodynamics. The tunnel effect also raises hopes of lowering the temperature of fusion carried out in a controlled manner. Thanks to the tunnelling phenomenon, the emission of $\alpha$ particles occurs in the radioactive decay of massive atomic nuclei.

    In modern technology, the operation of many semiconductor electronic components (e.g.\ the tunnel diode) and of devices such as the scanning tunnelling microscope is based on the tunnelling phenomenon.